Recruiting Chapter

### 2020-12-19 manuscript.Rmd
copy-edit section 3

We utilize network methods rather than conventional descriptive statistics for two, related reasons. First, social network analysis privileges the relationship between actors. The discussion of Stevens and Khan suggests that for private school students, "it is not just the quality of the students that gets them into college but the quality of the relationship between elite high schools and colleges" [@RN4407] [pg. 175]. Second, a visit from a college/university admissions counselor to a high school can be conceived as a "network tie" connecting two actors in a social network.

We briefly define essential social network analysis concepts and their operationalization in our analyses.  A social network consists of a set of actors -- referred to as "vertices" -- and the connections -- referred to as "edges" between these actors.  Whereas "one mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two mode" networks consit of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an an actor-movie network consists of actors -- mode 1 -- who appear in movies -- mode 2 and an actor shares an edge with a movie if the actor appears in the movie.  The social network analyzed in this chapter is a two-mode network, where vertices consists of colleges/universities (mode 1) and high schools (mode 2) and an "edge" is defined as a visit from a college/university to a private high school [DELETE THIS SENTENCE OR THE REDUNDANT ONE IN BELOW PARAGRAPH].

Our analyses draw from analyses of "corporate board-director" networks because these are the most commonly analyzed two-mode networks in the social networks literature [e.g.] [@RN4663]. Board-director networks consist of directors (mode 1) who serve on organizational boards (mode 2). Our school-college network is a two-mode network composed of two sets of vertices, private high schools (mode 1) and colleges/universities (mode 2). Off-campus recruiting visits are edges (e.g., high school $i$ shares an edge with college $j$ if high school $i$ receives at least one visit from college $j$) and visits can only occur between a school-college pair (not a school-school pair or a college-college pair). Our school-college network is a two-mode network composed of high schools (mode 1) who are visited by colleges and universities (mode 2). This (weighted) network is represented as a school-by-college matrix (e.g., a $500 \times 40$ matrix if our network contains 500 high schools and 40 colleges/universities) in which matrix cell $a_{i,j}$ identifies the number of visits that high school $i$ received from college/university $j$. [SHOW AN EXCERPT OF THE TWO MODE ADJACENY MATRIX, LIKE 4 BY 3 OR SOMETHING] [DEFINE WEIGHTED VS. UNWEIGHTED SOMEWHERE]

@@

Two-mode networks are often analyzed as one-mode networks [@RN2457]. @RN4663 states that "overlapping groups such as boards of directors form a two-mode membership network in which one can conceive of directors as nodes [vertices] connected by a tie of common board membership, or boards as nodes connected by a tie of one or more shared directors" [pg. xx]. Thus, a two-mode board-director network can be transformed into two one-mode networks. A weighted one-mode director network can be created as follows: create a director-by-director matrix (e.g., a $100 \times 100$ matrix if there are 100 directors in the network); two directors share an edge if they serve together on a board; matrix cell $a_{i,j}$ identifies the number of boards that include both director $i$ and director $j$ (e.g., $a_{i,j}=0$ if directors $i$ and $j$ do not serve on any of the same boards and $a_{i,j}=3$ if directors $i$ and $j$ do not serve on three boards together. Similarly, a one-mode board network is created based on a board-by-board matrix in which two boards share an edge if a director sits on both boards and matrix cell $a_{i,j}$ identifies the number of directors that serve on both board $i$ and board $j$. Following this approach, we create a one-mode high school network in which two private high schools share an edge if both receive a visit by the same college/university and cell $s_{i,j}$ of the school-by-school matrix $S$ indicates the number of colleges/universities that visited both school $i$ and school $j$ [SHOW EXCERPT OF ONE-MODE SCHOOL BY SCHOOL MATRIX, LIKE 3 BY 3 OR SOMETHING]. Similarly, we create a one-mode college/university network (herein the college network) in which two colleges share an edge if both colleges visited the same high school and cell $c_{i,j}$ of the college-by-college matrix $C$ indicates the number of private high schools that received at least one visit from both college $i$ and college $j$ [SHOW EXCERPT OF ONE-MODE COLLEGE BY COLLEGE MATRIX, LIKE 3 BY 3 OR SOMETHING].

@@

A limitation of our data collection is that we collected off-campus visit data only from a convenience sample of X colleges and universities, as described above. Thus, for each college and university in our sample, we know all the private high schools they visited (assuming no measurement error). However, for each private high school in our sample, we only know visits they received from the set of X colleges/universities in our sample. Given this limitation, it may be useful to think of our data collection as resulting in X "ego networks." An "ego" is a single "focal" node (e.g., The University of Notre Dame). An ego network consists "of a focal node ("ego") and the nodes to whom ego is directly connected to (these are called "alters") plus the ties, if any, among the alters" [from http://www.analytictech.com/networks/egonet.htm#:~:text=Ego%20networks%20consist%20of%20a,form%20The%20human%20social%20network ]. For example, the University of Notre Dame ego network consists of all private high schools that received at least one visit from Notre Dame. Additionally, for each high school visited by Notre Dame, the ego network may include all colleges/universities in our sample that also visited the high school.

When analyzing ego networks, researchers often investigate the extent of "homophily" versus "heterophily" [@RN1722]. Homophily is the idea that two vertices are more likely to be connected if they have similar characteristics (e.g., two people are more likely to be freinds on Facebook if they have similar political ideologies). Heterophily, the opposite of homophily, is when pairs of connected vertices have different characteristics from one another. Analyses investigate the extent to which network ties are homopholous or heteropholous, focusing on the vertex characteristics of

@@

Many social network analyses utilize data-driven "community detection" algorithms to categorize vertices into a small number of groups (or cluster) based purely on patterns of network ties. Generally, groups are chosen to maximize the number of within-group ties while minimizing ties between members of different groups. We used XYZ community detection approach [add footnote w/ details?] to allocate private colleges and universities [? and private high schools] to clusters.

>>>>

We utilize network methods rather than conventional descriptive statistics for two, related reasons. First, social network analysis privileges the relationship between actors. The discussion of Stevens and Khan suggests that for private school students, "it is not just the quality of the students that gets them into college but the quality of the relationship between elite high schools and colleges" [@RN4407, p. 175]. Second, a visit from a college/university admissions counselor to a high school can be conceived as a "network tie" connecting two actors in a social network.

We briefly define essential social network analysis concepts and their operationalization in our analyses.  A social network consists of a set of actors -- referred to as "vertices" -- and the connections -- referred to as "edges" between these actors.  Whereas "one-mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two-mode" networks consist of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an actor-movie network consists of actors (mode 1) who appear in movies (mode 2), and an actor shares an edge with a movie if the actor appears in the movie. <!--The social network analyzed in this chapter is a two-mode network, where vertices consists of colleges/universities (mode 1) and high schools (mode 2), and an edge is defined as a visit from a college/university to a private high school [DELETE THIS SENTENCE OR THE REDUNDANT ONE IN BELOW PARAGRAPH].-->

Our analyses draw from analyses of "corporate board-director" networks because these are the most commonly analyzed two-mode networks in the social networks literature [e.g., @RN4663]. Board-director networks consist of directors (mode 1) who serve on organizational boards (mode 2). In our school-college two-mode network, the vertices are private high schools (mode 1) and colleges/universities (mode 2), and an edge is defined as an off-campus recruiting visits made between the two (e.g., high school $i$ shares an edge with college $j$ if high school $i$ receives at least one visit from college $j$). Visits can only occur between a school-college pair (not a school-school pair or a college-college pair). This (weighted) network can be represented as a school-by-college matrix (e.g., a $500 \times 40$ matrix if our network contains 500 high schools and 40 colleges/universities) in which matrix cell $a_{i,j}$ identifies the number of visits that high school $i$ received from college/university $j$. <!--[SHOW AN EXCERPT OF THE TWO MODE ADJACENY MATRIX, LIKE 4 BY 3 OR SOMETHING] [DEFINE WEIGHTED VS. UNWEIGHTED SOMEWHERE]-->

@@

Two-mode networks are often analyzed as one-mode networks [@RN2457]. @RN4663 states that "overlapping groups such as boards of directors form a two-mode membership network in which one can conceive of directors as nodes [vertices] connected by a tie of common board membership, or boards as nodes connected by a tie of one or more shared directors" (p. XXX). Thus, a two-mode board-director network can be transformed into two one-mode networks. A weighted one-mode director network can be created as follows: create a director-by-director matrix (e.g., a $100 \times 100$ matrix if there are 100 directors in the network); two directors share an edge if they serve together on a board; matrix cell $a_{i,j}$ identifies the number of boards that include both director $i$ and director $j$ (e.g., $a_{i,j}=0$ if directors $i$ and $j$ do not serve on any of the same boards and $a_{i,j}=3$ if directors $i$ and $j$ serve on three boards together). Similarly, a one-mode board network is created based on a board-by-board matrix in which two boards share an edge if a director sits on both boards and matrix cell $a_{i,j}$ identifies the number of directors that serve on both board $i$ and board $j$. Following this approach, we create a one-mode high school network in which two private high schools share an edge if they both receive a visit by the same college/university. Cell $s_{i,j}$ of the school-by-school matrix $S$ indicates the number of colleges/universities that visited both school $i$ and school $j$<!-- [SHOW EXCERPT OF ONE-MODE SCHOOL BY SCHOOL MATRIX, LIKE 3 BY 3 OR SOMETHING]-->. Similarly, we create a one-mode college/university network (herein the college network) in which two colleges share an edge if both colleges visited the same high school. Cell $c_{i,j}$ of the college-by-college matrix $C$ indicates the number of private high schools that received at least one visit from both college $i$ and college $j$<!-- [SHOW EXCERPT OF ONE-MODE COLLEGE BY COLLEGE MATRIX, LIKE 3 BY 3 OR SOMETHING]-->.

@@

A limitation of our data collection and analysis is that it is based on a convenience sample of 42 colleges and universities, as described above. Thus, for each college and university in our sample, we know all the private high schools they visited (assuming no measurement error). However, for each private high school in our sample, we only know visits they received from the set of 42 colleges/universities in our sample. Given this limitation, it may be useful to think of our data collection as resulting in 42 "ego networks." An "ego" is a single "focal" node (e.g., The University of Notre Dame). An ego network consists "of a focal node ('ego') and the nodes to whom ego is directly connected to (these are called 'alters') plus the ties, if any, among the alters" (p. XXX)<!-- [from http://www.analytictech.com/networks/egonet.htm#:~:text=Ego%20networks%20consist%20of%20a,form%20The%20human%20social%20network ]-->. For example, the University of Notre Dame ego network consists of all private high schools that received at least one visit from Notre Dame. Additionally, for each high school visited by Notre Dame, the ego network may include all colleges/universities in our sample that also visited the high school.

When analyzing ego networks, researchers often investigate the extent of "homophily" versus "heterophily" [@RN1722]. Homophily is the idea that two vertices are more likely to be connected if they have similar characteristics (e.g., two people are more likely to be friends on Facebook if they have similar political ideologies). Heterophily, the opposite of homophily, is when pairs of connected vertices have different characteristics from one another. In our analyses, we investigate the extent to which ego network ties are homophilous or heterophilous, focusing on the vertex characteristics of geographic region, religious affiliation, racial composition, and academic ranking.

@@

Many social network analyses utilize data-driven "community detection" algorithms to categorize vertices into a small number of groups (or cluster) based purely on patterns of network ties. Generally, groups are chosen to maximize the number of within-group ties while minimizing ties between members of different groups. We performed hierarchical cluster analysis to allocate colleges and universities into clusters based on their private high schools visits.<!--We used XYZ community detection approach [add footnote w/ details?] to allocate private colleges and universities [? and private high schools] to clusters.-->
